\chapter{Preface}\label{sec:preface}

% what is this book about?
% TAPL has a paragraph describing at a high level what the book is about

% relational programming
% the miniKanren language
% lots of examples



% why am I writing this book?

% language has changed significantly since TRS.
% new constraint operators, and a framework for constraint logic programming cKanren.  also, nominal logic programming, tabling, etc.  These topics are too advanced to easily fit into a second edition of TRS, which in MHO should remain focused on traditional logic programming.  Could have The Constrained Schemer could be a little followup, though.



%% This book is about \emph{relational programming}.  Relational programming

%% why a term to distinguish relational programming from logic programming?

%% erasing the distinction between the inputs to a function, and the function's output

%% no longer distinguish between the inputs and output(s) of a function




%% %
%% Relational programming is concerned with writing programs as
%% %
%% The defining characteristic of relational programming is that

%% several themes [concerns?] will emerge, the most important of which are
%% \emph{divergence}, and \emph{negation} (which, we will see, are
%% closely related)

%% %



This book is about \emph{relational programming}.
%
Just as functional programs model mathematical functions, relational
programs model mathematical relations.
%
This relational view of programming erases the distinction between a
program's inputs and outputs.
%
The resulting code is more flexible
than the functional, imperative, or object-oriented equivalents, often
in delightfully surprising and enlightening ways.
%
For example, we can run a relational program \emph{backwards},
determining which ``input'' values correspond to a given ``output''
value.



% Relational programming is intimately related to both logic programming
% and relational algebra, the theory behind relational databases.
%

%
Relational
%
\marginnote{This means avoiding all ``extra-logical'' operators, such
  as Prolog's cut ({\tt !}), {\tt assert} and {\tt retract}, {\tt is},
  and {\tt copy\_term}.}
%
programming can be seen as an especially pure form of
traditional (Prolog-style) logic programming, in which we commit
completely to the ideal of writing programs as relations.
%
We are willing to trade efficiency and expressiveness in favor of
increased flexibility for the programs we \emph{can} express.
%
In order to express more interesting programs, and to regain
efficiency, we will augment our core relational language with an
ever-larger set of \emph{constraint operators}.
%
For
%
\marginnote{Pure Prolog is actually a special case of constraint logic
  programming, over the domain of trees (\emph{CLP(tree))}.}
%
this reason, our approach is closer to that of constraint logic
  programming than to traditional logic programming.
%
In constrast with other forms of logic programming, we will always
emphasize the \emph{relational} aspect of our programs.
%
Our motto for this book shall be:

\begin{quotation}
A program that doesn't run backwards isn't worth writing.\marginnote{With apologies to Alan Perlis.}
\end{quotation}


The
%
\marginnote{If Scheme is the best language for learning recursion, and Haskell for
learning about monads, then miniKanren is the best language for
understanding the Halting Problem.}
%
two main challenges we will face when writing programs as
relations are \emph{avoiding divergence} (looping forever) and
\emph{expressing negation}, which turn out to intimately connected.


%\marginnote{If you know nothing about logic programming, relational algebra, or databases, \emph{don't panic!} This book only assumes that you can write simple recursive programs in a functional language, and have a reading knowledge of Scheme.  See the \emph{Audience} section below for details.}



%
%Relational programs are unusually concise and flexible, often
%displaying behavior that surprises the programmer.
%
% Relational programming is arguably the ultimate approach to
% declarative programming, and may represent the next step in the
% evolution of high-level programming.

% I hope you'll find the relational style of programming as fascinating
% as I do, and will help push the statre-of-the-art in this important
% yet underused programming paradigm.

%% This book will teach you how to write relations that produce
%% interesting answers when running both forwards and backwards.  For
%% example, you will learn to write a relational interpreter for a subset
%% of Scheme: \mbox{\scheme|(evalo '((lambda (x) x) 5) val)|} associates
%% \scheme|val| with \schemeresult|5|. Of course, we can play more tricks
%% with \scheme|evalo|: \mbox{\scheme|(evalo exp '6)|} generates legal
%% Scheme expressions that \emph{evaluate} to 6, while
%% \marginnote{Douglas Hofstadter coined the term \emph{quine} to describe a program that evaluates to itself, in honor of logician Willard Van Orman Quine (1908--2000).  Writing quines has long been a favorite hacker activity, and quines are often featured in the The International Obfuscated C Code Contest (\url{http://www.ioccc.org/}). A delightful introduction to quines can be found in Doug's classic, \emph{GEB}: \cite{GEB79}} \mbox{\scheme|(evalo exp exp)|} generates Scheme expressions that evaluate to themselves.



%% This\marginnote{The Datalog family of languages is even more closely related to relational databases than the miniKanren language described in this book.  Datalog is more expressive than SQL, but less expressive than miniKanren or Prolog. Datalog is sub-Turing complete, and cannot express infinite computations, which is generally a desireable property for a query language. More on Datalog can be found in: \cite{Ceri:1989:Datalog}} book is about writing programs that ``run backwards.''
%% %
%% Such programs represent mathematical relations; we therefore call them \emph{relational programs}, or just \emph{relations}.
%% %
%% Relational programming is closely related to relational databases, and can also be seen as an especially pure form of logic programming.
%% %
%% The perspective of this book is somewhat different, and is based on generalizing functions to relations. In other words, this book borrows more from functional programming than from logic programming or relational databases.


%% For writing our relational programs, we use the \emph{miniKanren} family of languages.  miniKanren\marginnote{Information on miniKanren, including links to implementations, can be found at the official miniKanren website: \url{http://minikanren.org/}} has been implemented in Scheme, Racket, Clojure (in the form of David Nolen's excellent {\tt core.logic} library), Ruby, Python, and several other ``host'' languages.  miniKanren is designed specifically for relational programming, and makes relational programming easy.

%% Okay, that last part is a lie.  Writing relational programs is $\ldots$ tricky, hence this book.  miniKanren does make relational programming \emph{less difficult} than it would be otherwise.  What \emph{is} true is that writing relational programs is extremely interesting, challenging, and fun.  As in, ``I can write a short, non-obfuscated program that will melt your brain,'' level of fun.


\section{Audience}\label{sec:preface:audience}

\marginnote{\begin{quote}I have attempted to deliver [these lectures] in a spirit that should be recommended to all students embarking on the writing of their PhD theses: imagine that you are explaining your ideas to your former smart, but ignorant, self, at the beginning of your studies!\end{quote}\begin{flushright}\vspace{-1em}---Richard P. Feynman\\\emph{The Feynman Lectures on Computation}\end{flushright}}
%
This book is written for
intermediate-to-advanced programmers, computer science students, and
researchers.  For this book, \emph{intermediate} means that you are
comfortable writing simple recursive procedures in a functional
programming language, such as Scheme, Racket, Clojure, Lisp, ML, or
Haskell.  I also assume you have a reading knowledge of Scheme.
%
No knowledge of relational programming, logic programming, or programming
language theory is required.

If you want to learn about relational programming, but are new to
programming, Dan Friedman, Oleg Kiselyov, and I have written a book
just for you, called \emph{The Reasoned Schemer}\cite{trs}. In that
book we assume you are familiar with the material in \emph{The Little
  Schemer}\cite{Friedman:1996:LS:230223}, which is a very gentle
introduction to recursion and functional programming.

You might also benefit from \emph{The Little Schemer} if you are an
experienced programmer but feel uncomfortable with recursion.  More
traditional texts on functional programming in Scheme include
\emph{Scheme and the Art of Programming}\cite{Springer:1989:Art}, the
first edition of \emph{How to Design
  Programs}\cite{Felleisen:2001:htdp1}\marginnote{(full text at
  \url{http://htdp.org/})}, and the classic \emph{Structure and
  Interpretation of Computer
  Programs}\cite{Abelson:1996:SIC:547755}\marginnote{(full text at
  \url{http://mitpress.mit.edu/sicp/full-text/book/book.html})}.

If you are an experienced functional programmer, but do not know
 Scheme, the beginning of \emph{Structure and Interpretation
  of Computer Programs} should get you up to speed, while \emph{The Scheme Programming Language, Fourth
  Edition}\cite{Dybvig:2009:SPL:1618542}\marginnote{(full text at
  \url{http://www.scheme.com/tspl4/})} describes the language in detail.



%% I have chosen Racket\marginnote{\url{http://racket-lang.org/}} as the host
%% language for this book, and cKanren as the default miniKanren
%% implementation, for several important reasons.
%% %
%% Racket's cKanren library currently has the most advanced experimental
%% extensions to miniKanren; we will need all of these features for the
%% relations we are going to write.
%% %
%% Racket is distributed under a Free
%% Software\marginnote{\url{http://www.gnu.org/licenses/old-licenses/lgpl-2.1.html\#SEC1}}
%% license, which is in the open spirit of this book.
%% %
%% Racket also has excellent documentation and debugging tools, and is
%% simple to download, install, and run.
%% %
%% The subset of Racket used in the book is essentially identical to
%% $R^{5}RS$\marginnote{\url{http://www.schemers.org/Documents/Standards/R5RS/}}
%% Scheme, the language used in previous academic work on miniKanren,
%% including my dissertation.
%% %
%% With the exception of a few Schemely features (such as
%% \scheme|call/cc|) that I explain where they first appear, this subset
%% of Racket should be easily readable to any functional programmer.

%% If you are a Clojurer and prefer to follow along in {\tt core.logic},
%% the first appendix describes the main differences between the cKanren
%% and {\tt core.logic} implementations of miniKanren.
%% %
%% It should be straight-forward to translate most relational programs
%% from cKaren to {\tt core.logic}.
%% %
%% Translating the implementations of miniKanren from Racket to Clojure
%% is another story.
%% %
%% I suggest you follow along with the Racket code, and then listen to
%% one of David Nolen's outstanding talks on the implementation of {\tt
%%   core.logic}.



% github

%\newpage

\section{The Language}\label{sec:preface:language}

The relations in this book are written in \emph{miniKanren}, a language
designed specifically for relational programming.
%
Other languages that appear well-suited to relational programming
include the logic programming languages
\lambdaProlog\cite{Miller:2012:PHL} and
\Godel\cite{Hill:1994:GodelLang}, and the Twelf specification and
proof language\marginnote{Twelf website: \url{http://twelf.org/}}.

miniKanren was first implemented in Scheme; over the past eight years
miniKanren has been implemented in Racket,\marginnote{Links to these
  implementations, and many other miniKanren resources, can be found
  at the official miniKanren website: \url{http://minikanren.org/}}
Clojure, Ruby, Python, and many other ``host'' languages.
%
There are dozens of experimental miniKanren implementations in Scheme
alone, and several other host languages have more than one miniKanren
implementation.
%
Like Lisp and Scheme, miniKanren is really a \emph{family} of related
languages.
%

This diversity of implementations and host languages has its advantages.
%
Interest in miniKanren has largely been driven by the popularity of
David Nolen's excellent {\tt core.logic}\marginnote{{\tt core.logic}:\\\noindent\url{https://github.com/clojure/core.logic}} Clojure library.
%
Claire Alvis continues to add advanced constraint solving features to the
cKanren\marginnote{cKanren in Racket:\\\noindent\url{https://github.com/calvis/cKanren}}
(``constraint Kanren'') Racket library.
%
Other researchers and hackers are experimenting with their own
language extensions and optimizations, greatly accelerating the
evolution of miniKanren.
%
Unfortunately, this diversity poses a dilemma for both the author
and readers of this book: which host language and implementation to
use?

% I intend to create separate Scheme, Racket, and Clojure editions of
% this book.
%
% Since Scheme is the language I know best, and since all academic work
% on miniKanren has used Scheme as the host language, I have decided to
% write the Scheme edition first.

Since Scheme is the language I know best, and since most academic work
on miniKanren has used Scheme as the host language, I have decided to
write this book using Scheme.  Anyone interested in ``porting'' the
book to Racket, Clojure, or another host language may do so with my
blessing.

%% In order to treat all my readers as first-class citizens, I have
%% decided to write three editions of this book, one for each of the main
%% host languages and implementations: Scheme/miniKanren, Racket/cKanren,\marginnote{Racket and Clojure editions of this book can be found at: \url{https://github.com/webyrd/relational-programming-in-miniKanren}}
%% and Clojure/{\tt core.logic}.
%% %
%% You are reading the Scheme version---you can find the Racket and
%% Clojure versions at the book's website.




\section{Goal}\label{sec:preface:goal}

% What is the goal of the book?  What do I want the reader to get
% out of the book, think about, or be able to do after reading?

%\marginnote{To some extent, this book can be thought of as an updated, greatly expanded, and much more accessible version of my dissertation (full text at \url{http://gradworks.umi.com/3380156.pdf}):\cite{byrdthesis}}

%The two high-level goals for this book are to {\bf make relational programming
%accessible} to a broader audience, and to {\bf describe the state-of-the-art}
%in the design, implementation, and use of the miniKanren language.
%

The goal of this book is to show how to effectively use miniKanren to
write relational programs.
%
%An important side-effect of these goals is to {\bf provide the background}
%needed to understand academic work on miniKanren.
%
To this end, the book presents a variety of non-trivial miniKanren
relations, written in what I consider to be idiomatic style, and shows
how these relations were derived.
%
Many of these examples draw from academic papers on miniKanren, and
focus on programming language theory (interpreters, type inferencers,
etc.).
%
Other examples, such as finite state machines, should be immediately
understandable by a wider audience of programmers.
%
To\marginnote{In this sense the book emulates Benjamin Pierce's outstanding \emph{Types and Programming Languages} (\emph{TAPL}), which is about types, but also contains excellent introductions to other programming languages related topics:\cite{tapl}} make the book as accessible as possible, all of these concepts are
explained either in the main text or in appendices.
%
This book can therefore be used as an informal introduction to programming language concepts.

%% The high-level goals for this book are to make relational programming
%% accessible to a broader audience, and to explain the state-of-the-art
%% in the design, implementation, and use of the miniKanren language.

%% Specifically, this book aims to:

%% \begin{enumerate}
%% \item describe the current state of the miniKanren language, and
%%   its Constraint Logic Programming (CLP) extensions;
%% \item present a variety of relational programs that exemplify the
%%   relational style, and that introduce important relational idioms;
%% \item explain how to derive miniKanren relations from pure Scheme
%%   functions in a systematic manner;
%% \item explain in detail the canonical implementation of core miniKanren,
%%   and show how to hack the implementation to change or extend the
%%   language;
%% \item explain how constraint solving works in the cKanren
%%   Constraint Logic Programming framework, and show how to design and
%%   implement new constraint domains;
%% \item demonstrate how to debug miniKanren relations;
%% \item explain how miniKanren differs from related languages,
%%   including Prolog, Curry, Mercury, and Oz;
%% \item describe important open problems in relational programming, and
%%   limitations of miniKanren;
%% \item explain the rationale behind key design decisions for miniKanren;
%% \item use relational programming to explain important and interesting
%%   concepts from programming languages, such as interpreters, type
%%   inferencers, and Continuation-Passing Style.
%% \end{enumerate}

%% \noindent
%% An important side-effect of these goals is to provide the background
%% needed to understand academic papers and talks on miniKanren.


%% One goal of this book is to present a variety of non-trivial
%% miniKanren relations, and to show how these relations were derived.
%% Many of these examples draw from academic papers on miniKanren, and
%% focus on programming languages theory (interpreters, type inferencers,
%% etc.).  Other examples, such as finite state machines, should be of
%% interest to a wider audience of programmers.


% PL-specific terms should be described,
% or at least the reader should be given very specific pointers to
% additional, accessible reading.

% Decribe the current state-of-the-art for relational programming in
% miniKanren.

% Show, in detail, how miniKanren relations can be derived from
% pure Scheme functions.


% Give examples of good relational style.

% Give a more detailed explanation of the miniKanren implementation than is found in TRS or our academic papers.  Show how to hack the implementation to change the behavior of miniKanren, or add new operators.

% Explain how the constraint solving works, and how to design and implement new constraint domains.

% Explain how to debug miniKanren programs.  lol

% Explain the difference between miniKanren and related languages, including Prolog, Curry, Mercury, and Oz.

% Describe the open problems in relational programming, and limitations of miniKanren.

% [Optional]
% Show how to ``play in the mud'' with non-relational features.

% [Optional]
% Explain design decisions made for miniKanren.

% [Optional]
% Give a history of miniKanren (with help from Dan and Oleg)

% [Optional]
% Describe real-world uses of miniKanren.

%\newpage

\section{Running the Code}\label{sec:preface:code}

All of the source code in the book can be found at:

\begin{fullwidth}
\noindent
\url{https://github.com/webyrd/relational-programming-in-miniKanren/tree/master/src}
\end{fullwidth}


\noindent 
All code has been tested under Vicare Scheme,\marginnote{Vicare Scheme, developed by Marco Maggi, is a fork of my friend Abdulaziz Ghuloum's Ikarus Scheme: \url{http://marcomaggi.github.io/vicare.html}} and should run with little or no modification under other implementations that are compatible with the \RfiveRSsp or \RsixRSsp\marginnote{\RfiveRSsp Standard: \url{http://www.schemers.org/Documents/Standards/R5RS/}\\\noindent\RsixRSsp Standard: \url{http://www.r6rs.org/}} Scheme standards.




\section{Margin Notes}\label{sec:preface:margin-notes}
\marginnote{This book is set using the ``Tufte-Style Book''
  \LaTeX\ style, freely available from
  \mbox{\url{http://www.LaTeXTemplates.com}}}

This book is typeset in the style of Edward Tufte's
magnificent and beautiful \emph{The Visual Display of
  Quantitative Information}\cite{Tufte:1986:VDQ:33404}.  I share
Tufte's love of margin notes,\marginnote{Another great lover of
both margin notes and footnotes was David Foster Wallace (1962--2008).
%
DFW's mastery of notes is best demonstrated by his essay,
``Host,'' in:\cite{dfw:lobster:2005}}
\marginnote{Wallace also loved \emph{marginalia};
 the \emph{Harry Ransom Center}'s DFW collection includes heavily annotated books
  from Wallace's personal library:
  \url{http://www.hrc.utexas.edu/press/releases/2010/dfw/books/}.}
and use them in this book to help solve the problem of addressing readers with
widely varying knowledge of computer science and programming.
%
To make the book as accessible as possible, in the main text I assume
the reader is the hypothetical intermediate-level programmer or
student described in the \emph{Audience} section above.
%
In the margin notes, however, anything goes, and I reserve the right
to geek out whenever necessary (or unnecessary).


% \section{Exercises}\label{sec:preface:exercises}


% \section{Typographic Conventions}\label{sec:preface:typographic-conventions}

% superscript letters: o, e
% == entered as ==


 \section{Acknowledgements}\label{sec:preface:acknowledgements}




%   Targeted level of theoretical CS knowledge for existing miniKanren resources
% <----------------------------------------------------------------------------->
% non-existent   basic            intermediate     Undergrad     Graduate     PhD
%                programming      hobbyist         CS            CS           in
%                literacy                          degree        degree       PL
%
%
%              |----------- The Reasoned Schemer -----------|
%
%                                             |---- Weekly Google Hangouts ----|
%
%                                                 |--- My PhD dissertation ----|
%
%                                                 |----- Clojure/conj,
%                                                        Clojure/West,
%                                                        Strange Loop, 
%                                                        Flatmap talks --------|
%
%                                                      |--- Academic Papers ---|
%
%
%                                      |------------ *this book* --------------|
%                                      |--- main text -----|-- margin notes ---|

% Rules: try to target intermediate-to-advanced hobbyists, working
% programmers, and undergrad students in main text.  Allowed to geek
% out in margin notes: anything can go!  Intro chapters and appendices
% will help bring intermediate reader up to speed on CS (and,
% especially, PL-related) topics, such as interpreters, continuations,
% type theory, and program transformations.  This shouldn't be too
% boring, though, since can followup these chapters with the same
% ideas demonstrated relationally (for example, relational
% interpreters, relational CSPer, etc.)  Advanced, optional chapters
% and appendices can explore topics of arbitrary complexity.



%%% SPJ style: describe contributions first.  Then, write to satisfy
%%% the contributions.  I'm done writing once I've backed up all of
%%% the claimed contributions.




%%% I should make sure I have plenty of examples of relational
%%% programming not drawn from PL theory, to make the book more
%%% accessible.  Automata and regex are good examples.  Need to look
%%% for more.  Can borrow some ideas from Prolog books.


% Critical questions that should be answered in the preface:


% Who is the intended audience?  (Always keep this in mind...)  For
% the Little books, the intended reader is a bright but ignorant high
% school student.

% What is the required background of the reader?  Tied closely to the
% first question, obviously.





% quote?

% what is the origin of this book?
% Talk about TRS, thesis, papers, talks, C311/B521

% Goals for the book/what the reader should be able to do after reading the book

% point of view of the book

% language choice/host language/variants of miniKanren

% special instructions to readers?
% what should the reader know?
% typographic conventions?
% how does this book fit in with TRS and my dissertation?
% what parts of this work have previously appeared, and where?

% where can the reader find the code (and the latest version of the
% book)?  How would the reader run the code?

% structure of the book


% Acknowledgements

%% mK developers/co-authors from previous works

%% C311/B521 students

%% IU grad students

%% Google Hangout participants

%% Clojure community

%% creators of Scheme

%% pioneers in logic programming


%% Structure of TAPL preface:
% One para super-brief description of topic of book.
% * \section{Audience}
% * \section{Goals}
% \section{Structure} % (with chapter dependency figure)
% \section{Required Background}
% \section{Course Outlines}
% \section{Exercises}
% \section{Typographic Conventions}
% \section{Electronic Resources}
% \section{Acknowledgements}
% Quotes at the end

\wspace

\noindent
William E. Byrd\\
\noindent
Salt Lake City, Utah\\
\noindent
\monthyear
